### 2024-10-23 teaching/math240f24/draft.tex
administrative details
lecture 22 oct

For the other direction, we once again
let $a = nq_1 + r_1$ and $b = nq_2 + r_2$ from the division algorithm.
Now the assumption is that $r_1 = a\percent n = b\percent n = r_2$. So
$$a - b = (q_1-q_2) n + (r_1 - r_2) = (q_1-q_2)n.$$
This shows that $a\equiv b\pmod n$.\slug
\bye

\noindent\datestamp{14}{} \hskip\parindent
This proposition is useful in practice. For example, suppose we want to know whether
$22\equiv 8\pmod 3$. We calculate $22 = 7\cdot 3 + 1$, so $22\percent 3 = 1$; meanwhile
$8 = 2\cdot 3 + 2$, so $8\percent 3 = 2$. By the proposition, this means that
$22\not\equiv 8\pmod 3$.

@@

\proclaim Proposition \advthm. Let $a,n\in \ZZ$ with $n\ge 2$. Then if $ab \equiv 1\pmod n$
and $ac\equiv 1\pmod n$, then $b\equiv c\pmod n$.

@@

\proclaim Theorem \advthm. Let $a,n\in \ZZ$ with $n\ge 2$. Then
\medskip
\item{i)} $a$ is invertible modulo $n$ if and only if $\gcd(a,n) = 1$; and
\smallskip
\item{ii)} if $a$ is invertible, then there is a unique integer $b\in [0,n-1]$ such that
$ab\equiv 1\pmod n$. Namely, if
$$1 = sa + tn$$,
then we can set $b = s\percent n$.
\medskip

@@

For part (ii), suppose that $s,t\in \ZZ$ are such that $1 = sa+tn$. Then
taking this equation modulo $n$, we have
$$ 1 = sa+tn \equiv_n sa + 0t \equiv_n sa.$$
Letting $b = s\percent n$, we have $1 = ba$, since $s$ and $b$ belong to the same
equivalence class modulo $n$, and $b$ is an integer in $[0,n-1]$ by the division algorithm.
This proves existence of such a $b$. To show uniqueness, suppose that $b$ and $c$ are both
integers in $[0,n-1]$ with $ab \equiv 1\pmod n$ and $ac\equiv 1\pmod n$. Then
$ab\equiv ac\pmod n$, and multiplying on both sides by $b$, we have $bab\equiv bac\pmod n$.
But $ba\equiv 1\pmod n$, so cancelling this out in the equation, we have $b\equiv c\pmod n$.
In other words, $n\divides b-c$. But since $b,c\in [0,n-1]$, we have $b-c\in [-n+1, n-1]$,
and the only divisor of $n$ in this range is $0$. Hence $b-c = 0$ and we conclude that $b=c$.\slug

@@

\noindent\datestamp{15}{} \hskip\parindent
We now investigate what happens in the particular case where we the modulus $n$ is prime.
In this situation we have the following proposition.

@@

@@

>>>>

For the other direction, we once again
let $a = nq_1 + r_1$ and $b = nq_2 + r_2$ from the division algorithm.
Now the assumption is that $r_1 = a\percent n = b\percent n = r_2$. So
$$a - b = (q_1-q_2) n + (r_1 - r_2) = (q_1-q_2)n.$$
This shows that $a\equiv b\pmod n$.\slug

\noindent\datestamp{22}{x} \hskip\parindent
This proposition is useful in practice. For example, suppose we want to know whether
$22\equiv 8\pmod 3$. We calculate $22 = 7\cdot 3 + 1$, so $22\percent 3 = 1$; meanwhile
$8 = 2\cdot 3 + 2$, so $8\percent 3 = 2$. By the proposition, this means that
$22\not\equiv 8\pmod 3$.

@@

\edef\propinversesunique{\the\thmcount}
\proclaim Proposition \advthm. Let $a,n\in \ZZ$ with $n\ge 2$. Then if $ab \equiv 1\pmod n$
and $ac\equiv 1\pmod n$, then $b\equiv c\pmod n$.

@@

\edef\thminvertiblemodn{\the\thmcount}
\proclaim Theorem \advthm. Let $a,n\in \ZZ$ with $n\ge 2$. Then
\medskip
\item{i)} $a$ is invertible modulo $n$ if and only if $\gcd(a,n) = 1$; and
\smallskip
\item{ii)} if $a$ is invertible, then there is a unique integer $b\in [0,n-1]$ such that
$ab\equiv 1\pmod n$. Namely, if
$$1 = sa + tn,$$
then we can set $b = s\percent n$.
\medskip

@@

For part (ii), suppose that $s,t\in \ZZ$ are such that $1 = sa+tn$. Then
taking this equation modulo $n$, we have
$$ 1 = sa+tn \equiv_n sa + 0t \equiv_n sa.$$
Letting $b = s\percent n$, we have $1 = ba$, since $s$ and $b$ belong to the same
equivalence class modulo $n$, and $b$ is an integer in $[0,n-1]$ by the division algorithm.
This proves existence of such a $b$. To show uniqueness, we suppose that $b$ and $c$ are both
elements of $[0,n-1]$ with $ab\equiv 1 \pmod n$ and $ac\equiv 1\pmod n$. By
Proposition~{\propinversesunique}, we have $b\equiv c\pmod n$;
in other words, $n\divides b-c$. But since $b,c\in [0,n-1]$, we have $b-c\in [-n+1, n-1]$,
and the only divisor of $n$ in this range is $0$. Hence $b-c = 0$ and we conclude that $b=c$.\slug

@@

We now investigate what happens in the particular case where we the modulus $n$ is prime.
In this situation we have the following proposition.

@@

\bye

@@

\medskip\boldlabel Computing large powers modulo {\mathbold p}.
The next application is not a precise mathemathical statement; rather, it's an observation
that Fermat's little theorem can allow us to compute what large powers are modulo a prime
without having to compute the number itself. We'll illustrate this by two examples.

First we compute $25^{134} \pmod{11}$. We begin by noting that
$$25\equiv 3\pmod{11},$$
so what we really want is $3^{134}\pmod{11}$. By Fermat's little theorem, $3^{10}\equiv 1\pmod{11}$,
so we may cast out multiples of $10$ from $134$ to arrive at the easy computation
$$3^{134} \equiv 3^{10\cdot 13+4} \equiv (3^{10})^{13}\cdot 3^4 \equiv 3^2 \cdot 3^2
\equiv 9\cdot 9\equiv (-2)(-2) \equiv 4\pmod{11}.$$

Now let's try $25{134} \pmod{14}$. Since $14$ is not prime, we can no longer use Fermat's little theorem.
The trick we shall use is to square $25$ repeatedly. First note that $25\equiv -3 \pmod{14}$,
so $25^2 \equiv 9\pmod{14}$. Then, squaring again, we get
$$25^4 \equiv 9^2 \equiv (-5)^2 \equiv 25\equiv -3\pmod{14}.$$
This means that repeated squaring of $25$ cycles between $-3$ and $9$ modulo $14$, so we immediately
have
$$25^8\equiv 9,\qquad 25^{16} \equiv -3,\qquad 25^{32}\equiv 9,\qquad 25^{64} \equiv -3,$$
and
$$25^{128}\equiv 9$$
modulo $14$. Now writing the exponent $134$ as a sum of powers of two allows us to conclude that
$$ 25^{134} = 25^{128+4+2} = 25^{128}\cdot 25^4 \cdot 25^2 \equiv 9\cdot (-3)\cdot 9 \equiv (-3)(-3)
\equiv 9\pmod{14}.$$

\medskip\boldlabel Fermat's primality test.
Fermat's little theorem says that if $n$ is prime, then for all integers $a$ with $a\not\equiv 0\pmod n$
one has $a^{n-1}\equiv 1\pmod n$. The contrapositive of this statement is that if there
is some integer $a$ with $a\not\equiv 0\pmod n$ and $a^{n-1}\not\equiv 1\pmod n$, then $n$ is not prime.
This leads us to the following algorithm that can (usually) tell us when a number is not prime.

\algbegin Algorithm F (Fermat's primality test). Given an integer $n\ge 4$,
this algorithm will attempt to declare that
$n$ is not prime. If the algorithm fails to do so, then the test is inconclusive ($n$ may either
be prime or not).
\algstep F1. [Initialise.] Set $a \gets 2$. (We can skip the step $a=1$ since $a^{n-1}$ will
always equal $1$ in this case.)
\algstep F2. [Compute power.] Set $b\gets a^{n-1} \% n$. (So $0\le b< n$.)
\algstep F3. [Not prime?] If $b \neq 1$, output ``Not prime,'' and terminate the algorithm.
\algstep F4. [Loop?] Set $a\gets a+1$. If $a=n$, output ``Inconclusive,'' and terminate the algorithm.
Otherwise, go to step F2.\slug

Let's see a couple of examples. Suppose we want to find out if $9$ is prime or not (we both know it isn't
but play along for a second). In the very first step loop, when $a=2$, we have
$$2^{9-1} = 2^8 = (2^4)^2 = 16^2 \equiv (-2)^2 = 4\pmod 9,$$
so since $4 \not\equiv 1\pmod 9$, we conclude right away that $9$ is not prime.

A harder one. Is $341$ prime? We start with $a=2$. Noting that $3\cdot 341 = 1023$, we have
$$2^{340} = (2^{10})^{34} = 1024^{34} \equiv 1^{34}\equiv 1\pmod{341},$$
so the test is inconclusive after the first loop. Fine. If we try $a=3$ next, we will
find that $3^{340}\equiv 56\not\equiv 1\pmod{341}$, so we conclude that $341$ is not prime.
(Those wishing to practise computing large powers modulo $n$ might like to verify the fact
that $3^{340}\equiv 56\pmod {341}$. Use the squaring trick.)

If $n$ causes Algorithm~F to output ``Inconclusive,'' it is very likely that $n$ is prime. However,
there are rare cases of composite numbers $n$ that cause Algorithm~F to output ``Inconclusive.''
These are called {\it Carmichael numbers}. The smallest one is $561$.

\medskip\boldlabel Vernam's one-time pad.
Alice and Bob want to exchange secret messages over a public channel without anyone
being able to decipher them. A message can be encoded as a binary string $M$.
Here is one way that Alice and Bob can securely send encrypted messages to each other.
It is called Vernam's one-time pad, named for G.~Vernam, an {\mc AT\&T} engineer who patented
the method in 1917.

First, they need to meet in person and generate a long sequence of random bits. (How can
one generate random numbers? This is an interesting quasi-philosophical question.) Each of
them keeps a copy of this sequence, call it $S$.

Now suppose Alice and Bob separate and Alice wants to send the message $M = 0110$ to Bob. Of course,
she shouldn't send this string directly, as it could be intercepted by foes. She needs to use
the sequence $S$. Suppose
the first four bits of $S$ are $1101$. To encrypt her message, Alice adds each bit of $M$
to the corresponding bit of $S$, modulo $2$. We will denote this by the $\oplus$ operation.
The encrypted message is $M'=0110\oplus 1101 = 1011$. She sends it to Bob.

Bob receives the string $M' = 1011$. To decode it and recover the message $M$, Bob simply has to
perform the $\oplus$ operation using the (same) first four bits of $S$ that Alice used
to encode the message in the first place. Indeed, we see that $1101\oplus 1101 = 0110 = M$.
The reason that this works is that for any binary string $S$, we have $S\oplus S = 0$,
so performing the $\oplus$ operation twice always returns the original string.

If they want to send more messages, they simply discard the bits they used before and continue
using the next few bits of $S$.

The advantage to this approach is that an adversarial third-party, reading $M$', gains
no information about the original string $M$, provided they do not know the sequence $S$.
In other words, it is entirely secure.
The disadvantages of Vernam's one-time pad are that Alice and Bob must meet in person beforehand,
they need to find a way to generate random bits securely, and they need to periodically
repeat this process to replenish the bits once they run out.

\medskip\boldlabel The RSA cryptosystem.
Now we move on to a more sophisticated method of encrypting and decrypting messages.
It is called the {\it {\mc RSA} cryptosystem}, named for R.~Rivest, A.~Shamir, and L.~Adleman,
who described the algorithm in 1977.

Here's the scenario. Alice wants people to be able to send her messages such that
only she can decode them. To do so, she needs to create a {\it public key}, which allows
people to encrypt their messages before sending them to her, and then a {\it private key}
that will allow her to decrypt messages. (Only she knows the private key.)

To set things up, Alice needs two large primes, call them $p$ and $q$. (Something like $200$
bits is sufficient.) Then she computes $n=pq$. Now, she picks some integer $k$ with $\gcd(k,n) = 1$.
By Theorem~{\thminvertiblemodn}, there is some integer $s$ such that $ks=1$. Alice publishes
the integer $n$ and the public key $k$. The private key is $s$; this she keeps to herself.

Now Bob wants to send a secret message $M$ to Alice. Considering $M$
as an integer (in binary), a requirement for the algorithm to work is $M<n$. Bob computes
$\bar M = M^k\% n$ and sends $\bar M$ to $A$. (We have $0\le \bar M < n$ by the division
algorithm.)

Alice receives $\bar M$. She computes ${\bar M}^s$. It turns out that this is $M$.
To prove this, we need a lemma.

\proclaim Lemma \advthm. Let $a$ and $b$ be integers.
\medskip
\item{i)} For all positive integers $m$ and $n$,
if $a\equiv b\pmod{mn}$, then $a\equiv b\pmod m$ and $a\equiv b\pmod n$.
\medskip
In the case that $m$ and $n$ are different primes, the converse holds; that is, the following
holds.
\medskip
\item{ii)} For all primes $p$ and $q$ with $p\ne q$, if $a\equiv b\pmod p$
and $a\equiv b\pmod q$, then $a\equiv b\pmod{pq}$.
\medskip

\proof Assume that $a\equiv b\pmod{mn}$, so there exists $k\in \ZZ$ such that
$a-b = kmn$. This means that $a-b\equiv 0\pmod m$ and $a-b\equiv 0\pmod n$,
so $a\equiv b\pmod m$ and $a\equiv b\pmod n$.

To prove (ii), suppose that there exist $k,l\in \ZZ$ such that $a-b = kp$ and $a-b = lq$.
We have $kp=lq$, so $p\divides lq$. Because $p$ is prime, $p$ must divide either $l$ or $q$.
But $q$ is prime as well, so the only divisors of $q$ are $1$ and $q$. Hence $p\divides l$,
and we may write $l = rp$ for some $r\in \ZZ$. So we have
$$ a-b = lq = rpq,$$
implying that $a-b\equiv 0\pmod{pq}$; that is, $a\equiv b\pmod{pq}$.\slug

Now we shall prove that the encrypted message can be decrypted into the original message.

\parenproclaim Theorem {\advthm} ({\mc RSA} encryption). Let $p$ and $q$ be distinct primes and
let $n=pq$.

\bye
